Inferring the existence and direction of causal associations in the face of measurement error

### 2017-01-25 docs/manuscript.Rmd
updates

A common approach for inferring causal relationships between two traits is to use genetic instruments in mediation-based analyses. Here we show that when measurement error is present, commonly-used tools predicated on this approach are susceptible to making erroneous inference of both the existence of causality and the causal direction, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. This extension is less susceptible to measurement error than mediation-based analyses. We use it to demonstrate empirically that inferring the direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms. Other 'omic platforms are likely to be similarly susceptible, necessitating the use of sensitivity analyses to evaluate reliability.

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, potential pitfalls still persist and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

Non-differential measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more basic data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference. An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

@@

MR can operate on the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure. The causal effect of the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

>>>>

A common approach for inferring causal relationships between two traits is to use genetic instruments in mediation-based analyses. Here we show that when measurement error is present, widely used tools predicated on this approach, such as the causal inference test (CIT) are susceptible to making erroneous inference of both the existence of causality and the causal direction. Increasing sample size for these tests, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is relatively robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. This extension is less susceptible to measurement error than mediation-based analyses and only requires GWAS summary data. We use it to demonstrate empirically that inferring the direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms. Other 'omic platforms are likely to be similarly susceptible, necessitating the use of sensitivity analyses to evaluate reliability.

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, genetic instrumentation has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

Measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value. This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more basic data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

@@

MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure. The causal effect of the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.
